### 2025-01-08 doc/paper.tex
draw paper outline with bullet points and start related work section

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

\usepackage{tikz}
% Corporate Design of the University of Tübingen
% Primary Colors
\definecolor{TUred}{RGB}{165,30,55}
\definecolor{TUgold}{RGB}{180,160,105}
\definecolor{TUdark}{RGB}{50,65,75}
\definecolor{TUgray}{RGB}{175,179,183}

% Secondary Colors
\definecolor{TUdarkblue}{RGB}{65,90,140}
\definecolor{TUblue}{RGB}{0,105,170}
\definecolor{TUlightblue}{RGB}{80,170,200}
\definecolor{TUlightgreen}{RGB}{130,185,160}
\definecolor{TUgreen}{RGB}{125,165,75}
\definecolor{TUdarkgreen}{RGB}{50,110,30}
\definecolor{TUocre}{RGB}{200,80,60}
\definecolor{TUviolet}{RGB}{175,110,150}
\definecolor{TUmauve}{RGB}{180,160,150}
\definecolor{TUbeige}{RGB}{215,180,105}
\definecolor{TUorange}{RGB}{210,150,0}
\definecolor{TUbrown}{RGB}{145,105,70}

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2023} with \usepackage[nohyperref]{icml2023} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

\usepackage[accepted]{icml2023}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}
\usepackage{float}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Comparing different active learning query methods}

\begin{document}

\twocolumn[
\icmltitle{TODO: Funky active learning pun}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2023
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
	\icmlauthor{Lukas Weber}{first}
\end{icmlauthorlist}

\icmlaffiliation{first}{MSc Computer Science, Tübingen, GER}
\icmlcorrespondingauthor{Lukas Weber}{lukas2.weber@student.uni-tuebingen.de}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, Active Learning}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
	A lot of unlabelled data generally available but annotation is costly. A goal in fields where labelled data is not available  is to maximize performance with minimum amount of labeled data instances. \\
	Active learning selects instances for labelling based on a selection strategy. 
	\begin{itemize}
		\item Objective: Compare active learning query methods using modAL.
		\item Methods: Experiments with Logistic Regression (MNIST) and a CNN (CIFAR).
		\item Experiments: Vary initial dataset sizes (small, moderate, large).
		\item Key findings: Highlight main conclusions about performance in different scenarios.
	\end{itemize}
\end{abstract}

\section{Introduction}\label{sec:intro}
Start with a broader context like: Active learning aims to minimize the labeling cost by selecting the most informative samples for annotation by an oracle. However, the choice of an appropriate query strategy significantly impacts the effectiveness of active learning.

Motivation (Prolem i want to solve and why is the problem important)
\begin{itemize}
	\item Labelling expensive task
	\item important to reduce labelled data requirement in machine learning to broaded to be able to use it in data sparse environments
	\item role of  active learning in reducing annotation costs
	\item significance of evaluating query strategies
\end{itemize}
Define Active Learning and why query strategies are important. FOCUS IS ON COMPARING STRATEGIES UNDER DIFFERENT DATASET SIZES AND MODEL COMPLEXITIES\\ 
Objective
\begin{itemize}
	\item Compare different active learning strategies using the modAL framework
\end{itemize}

\textbf{Related Work: }What did the research do? prior comparisons of query strategies and highlight the gap that i am adressing
\begin{itemize}
	\item Previous works have explored strategies such as ..., .., .... While they have been widely tested, most evaluations focus on  single dataset scenarios or specific models
	\item Then the gaps: However, few studies analyze the impact of initial dataset size on query strategy performance, and even fewer investigate their behavior on simple versus complex models.
\end{itemize}
Previous papers:
\begin{itemize}
	\item \cite{revisitinguncertaintybasedquerystrategies} Compares various uncertainty based query strategies in context of fine-tuning transformer models for text classification. 
	\\
	Does not compare across different sizes of the pool, and for simple models.
	\item \cite{benchmarkingofquerystrategies} TOWARDS FUTURE
	DEEP ACTIVE LEARNING": Most papers concentrate on cifar or mnist, this paper evaluates six different datasets, including medical and visual inspection images. Also tackles the problem that experimental settings are not standardized, making evaluation of existing methods difficult. Also did a verification experiment where fully trained models are used to select most informative samples according to various query strategies and construct a labeled dataset. This is done to isolate and evaluate the effectiveness of query strategies indepently of training process/ underfitting or randomness due to early-stage training.

	\item \cite{comparativesurveydeepactive}: compares different query methods on mnist and cifar with a cnn (resnet18). Does not compare impact of different initial train set size and strategies on shallow classifiers.

	\item \cite{comparableactivelearning} highlight critical challenges in active learning research, including the lack of reproducible experiments and fair comparisons across domains. They propose a benchmark framework evaluating active learning strategies on tabular, image, and text datasets using robust evaluation protocols to mitigate variance and ensure comparability. While their work focuses on a wide range of domais, this paper only compares methods in the image domain. Furthermore, this paper focuses on the image domain. Specifically, it compares strategies for a simple logistic regression model and a CNN across different starting sizes of the annotated dataset. \\
	By examining different starting sizes, this work simulates various real-world scenarios: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
\end{itemize}

\textbf{Modal}: description and relevance for research maybe? The modAL framework(ref) has simplified the implementation of active learning pipelines, yet its comparative use across different strategies remains underexplored.

Prior studies
\begin{itemize}
	\item existing studies that compare active learning methods
	\item do they have gaps in the literature? (maybe: comparison under different dataset sizes, use with cnns, comparison of simple or complicated models)
\end{itemize}

\textbf{Contributions}
Comparative study of popular active learning query strategies using modAL framework. Contributions are as follows:
\begin{itemize}
	\item systemativ evaluation on two datasets: MNIST (Logistic Regression) and CIFAR (CNN)
	\item analysis across different scenarios: small, moderate and large initial datasets
	\item insight into performance impact of model complexity and initial dataset size on active learning performance
\end{itemize}

\textbf{Structure of the paper}
\begin{itemize}
	\item outline paper
\end{itemize}

\section{Methods}\label{sec:methods}
\label{data_and_methods}
\subsection{Datasets}
\begin{itemize}
	\item MNIST and CIFAR describe
	\item processing steps \\
	MNIST: Vectorized, flattened (describe input size), normalized
\end{itemize}

\subsection{Models}
\begin{itemize}
	\item Logistic Regression: Details on implementation and hyperparameters, how were they chosen?
	\item same for CNN
\end{itemize}

\subsection{Active Learning}
\begin{itemize}
	\item How are the splits (initial pool, initial training data) generated
	\item numbmer of iterations for active learning
	\item Used active learning query strategies
\end{itemize}

Metrics for evaluation (Accuracy, train error, confusion matrix, ...)

Tool: modAL, scikit

\section{Experiments}\label{sec:experiments}
Setup: details for reproducibility (random number generators)?
Random seed = 42
\\
Did I perform multiple runs?
\\

Scenarios
\begin{itemize}
	\item describe experiments and their settings and purposes
\end{itemize}

\section{Results}\label{sec:results}
Quantitative results: Tables or plots comparing the query strategies across scenarios. Also maybe key perforamnce differences (accuracy vs number of labeled examples)

\section{Discussion}\label{sec:discussion}
Qualitative analysis: Why certain strategies perform better or worse in specific settings. Trends between simple models and complex models
\\
\\
Impact of initial dataset size
\\
\\
Most methods require knowledge about the balance of the dataset and might be disadvantageous to use in case of a disbalanced dataset. In real world scenarios, this is not known. \\
Only image-based datasets were tested. Not vector- or text-based.

\section{Conclusion}\label{sec:conclusion}
Summary of findings and insights (maybe that a strategy works best with small datasets while another one is robust on bigger ones)
\\
Implications: practical recommendations?
\\
Future work? (More diverse datasets, exploring additional strategies, applying in real-world tasks)

\section{Cool ideas}
\begin{itemize}
	\item Show the typical framework of active learning in a figure in the introduction
\end{itemize}

\section{Good formulations}
From the active learning survey: Entropy is an information-theoretic
measure that represents the amount of information needed to “encode” a distri-
bution. As such, it is often thought of as a measure of uncertainty or impurity in machine learning.

\bibliography{bibliography}
\bibliographystyle{icml2023}

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2025-01-09 doc/paper.tex
Start with introduction and related works

@@
 \section{Introduction}\label{sec:intro}
-Start with a broader context like: Active learning aims to minimize the labeling cost by selecting the most informative samples for annotation by an oracle.
-However, the choice of an appropriate query strategy significantly impacts the effectiveness of active learning.
-
-Motivation (Prolem i want to solve and why is the problem important)
-\begin{itemize}
-	\item Labelling expensive task
-	\item important to reduce labelled data requirement in machine learning to broaded to be able to use it in data sparse environments
-	\item role of  active learning in reducing annotation costs
-	\item significance of evaluating query strategies
-\end{itemize}
-Define Active Learning and why query strategies are important. FOCUS IS ON COMPARING STRATEGIES UNDER DIFFERENT DATASET SIZES AND MODEL COMPLEXITIES\\ 
-Objective
-\begin{itemize}
-	\item Compare different active learning strategies using the modAL framework
-\end{itemize}
-
-\textbf{Related Work: }What did the research do? prior comparisons of query strategies and highlight the gap that i am adressing
+Complex tasks in supervised learning usually require large annotated datasets.
+However, the cost and time needed to label samples can be high, especially when expert knowledge in the according field is required.
+\\
+Active learning aims to reduce the overall labeling cost by selecting the most informative samples for annotation by an oracle and thus enable using machine learning for tasks where annotated data is sparse.
+\\
+The effectiveness of active learning relies heavily on the choice of query strategy, as it determines which samples are selected for annotation and thus directly influences model performance.
+In pool-based active learning, a model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data.
+Informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set.
+The model is retrained, and the process is repeated until a stopping criterion is met.
+\\
+In stream-based active learning, unlabeled data arrives sequentially.
+Thus, the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of the prediction and checking whether a threshold is reached.
+If not, the instance is discarded.
+\\
+This report focuses on comparing several pool-based as well as stream-based query strategies. \\
+\textbf{Related Work:} Several previous works explore different query methods for several tasks.
+\cite{revisitinguncertaintybasedquerystrategies} compares various uncertainty based query strategies in the context of fine-tuning transformer models for text classification.
 \begin{itemize}
@@
 \begin{itemize}
-	\item \cite{revisitinguncertaintybasedquerystrategies} Compares various uncertainty based query strategies in context of fine-tuning transformer models for text classification. 
-	\\
-	Does not compare across different sizes of the pool, and for simple models.
-	\item \cite{benchmarkingofquerystrategies} TOWARDS FUTURE
-	DEEP ACTIVE LEARNING": Most papers concentrate on cifar or mnist, this paper evaluates six different datasets, including medical and visual inspection images.
+	\item \cite{revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
+	\item \cite{benchmarkingofquerystrategies} Most papers concentrate on cifar or mnist, this paper evaluates six different datasets, including medical and visual inspection images.
 Also tackles the problem that experimental settings are not standardized, making evaluation of existing methods difficult.
@@
 Random seed = 42
+\cite{comparableactivelearning} mention reproducibility problems in models even when setting random number generators before each run.
+By training models multiple times in succession and checking whether the same configurations lead to different outcomes, it is shown that in this scenario, reproducibility is guaranteed.
+\textcolor{red}{TODO: Schnell ein skript aufsetzen und dann finale accuracy und so weiter vergleichen.}
 \\

### 2025-01-09 doc/paper.tex
Work on Introduction

@@
 \\
-This report focuses on comparing several pool-based as well as stream-based query strategies. \\
-\textbf{Related Work:} Several previous works explore different query methods for several tasks.
-\cite{revisitinguncertaintybasedquerystrategies} compares various uncertainty based query strategies in the context of fine-tuning transformer models for text classification.
-\begin{itemize}
-	\item Previous works have explored strategies such as ..., .., .... While they have been widely tested, most evaluations focus on  single dataset scenarios or specific models
-	\item Then the gaps: However, few studies analyze the impact of initial dataset size on query strategy performance, and even fewer investigate their behavior on simple versus complex models.
-\end{itemize}
-Previous papers:
-\begin{itemize}
-	\item \cite{revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
-	\item \cite{benchmarkingofquerystrategies} Most papers concentrate on cifar or mnist, this paper evaluates six different datasets, including medical and visual inspection images.
-Also tackles the problem that experimental settings are not standardized, making evaluation of existing methods difficult.
-Also did a verification experiment where fully trained models are used to select most informative samples according to various query strategies and construct a labeled dataset.
-This is done to isolate and evaluate the effectiveness of query strategies indepently of training process/ underfitting or randomness due to early-stage training.
-
-	\item \cite{comparativesurveydeepactive}: compares different query methods on mnist and cifar with a cnn (resnet18).
-Does not compare impact of different initial train set size and strategies on shallow classifiers.
-
-	\item \cite{comparableactivelearning} highlight critical challenges in active learning research, including the lack of reproducible experiments and fair comparisons across domains.
+This report focuses on comparing several pool-based as well as stream-based query strategies.
+The modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning} is used in the following experiments as it simplified the implementation of active learning pipelines, contains multiple already implemented querying strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
+\\
+Several previous works explore different query methods for different tasks. \\
+\cite{schröder_revisitinguncertaintybasedquerystrategies} compares various uncertainty based query strategies in the context of fine-tuning transformer models for text classification, while \cite{zhan_comparativesurveydeepactive} compares several query methods on MNIST and CIFAR.
+\\
+According to \cite{ueno_benchmarkingofquerystrategies} most papers concentrate on two image-based datasets: MNIST~\cite{mnist} and CIFAR~\cite{cifar}, while their paper evaluates strategies on six different datasets, including medical and visual inspection images.
+They also verify the effectiveness of query strategies independently of the training process, underfitting and randomness due to early-stage training in an experiment where fully trained models are used to select the most informative samples according to various query strategies and to construct a labeled dataset.
+\\
+\cite{werner_comparableactivelearning} agrees with their statement that experimental settings are not standardized across papers which makes comparisons of different existing and new query strategies difficult.
 They propose a benchmark framework evaluating active learning strategies on tabular, image, and text datasets using robust evaluation protocols to mitigate variance and ensure comparability.
-While their work focuses on a wide range of domais, this paper only compares methods in the image domain.
-Furthermore, this paper focuses on the image domain.
-Specifically, it compares strategies for a simple logistic regression model and a CNN across different starting sizes of the annotated dataset.
+\\
+While \cite{ueno_benchmarkingofquerystrategies} evaluates strategies across several image-based datasets from a wide range of image-domains like the medical and visual inspection images, \cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets.
+\\
+This report focuses on comparing strategies in the image domain.
+It compares query strategies for a simple logistic regression model and a more complex CNN to study the effect of model complexity on the optimal query strategy.
+\textcolor{red}{CITATION were able to show that [QUERY METHODS] work well on simple models but do not generalize well on more complex models.
+To guarantee real-world applicability, knowledge about this needs to be gained as model complexities vary across different tasks and datasets}.
+Furthermore, the impact of different sizes of the initial annotated dataset is additionally investigated which is something that few studies analyze in combination.
+The effect of the combination between model complexity and initial annotated dataset size is a research gap in previously mentioned papers.
 \\
 By examining different starting sizes, this work simulates various real-world scenarios: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
+% FROM HERE THIS IS NOT GOING TO BE PART OF THE FINAL PART
+Previous papers do not: Is everything here now mentioned previously? if yes, then it can be deleted.
+\begin{itemize}
+	\item \cite{schröder_revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
+	\item \cite{comparativesurveydeepactive}: Does not compare impact of different initial train set size and strategies on shallow classifiers.
+
+	\item According to \cite{ueno_benchmarkingofquerystrategies}: \textcolor{red}{check if true:} Does not compare different sizes of the initial starting sets as well as how simple and more complex perform with the query methods
+
+	\item \cite{werner_comparableactivelearning}: They focus on a wide range of domains, this paper only compares methods in the image domain.
+I compare strategies for a simple logistic regression model and a complex CNN across different starting sizes of the annotated dataset.
+\\
+\end{itemize}
+% TO HERE
+The contributions of this paper are as follows:
+\begin{itemize}
+	\item \textbf{Systematic evaluation of active learning query strategies} on MNIST with a logistic regression model and CIFAR-10 using a CNN to analyze their behavior across datasets and model complexities.
+	\item \textbf{Investigation of model complexity's influence} on the effectiveness of query strategies to provide knowledge into how shallow and deep models respond differently to active learning methods.
+	\item \textbf{Analysis of the initial labeled dataset size's impact} on the performance of query strategies to simulate real-world scenarios such as starting completely from scratch, having a small labeled dataset or fine-tuning a model.
 \end{itemize}
 
-\textbf{Modal}: description and relevance for research maybe?
-The modAL framework(ref) has simplified the implementation of active learning pipelines, yet its comparative use across different strategies remains underexplored.
-
-Prior studies
-\begin{itemize}
-	\item existing studies that compare active learning methods
-	\item do they have gaps in the literature? (maybe: comparison under different dataset sizes, use with cnns, comparison of simple or complicated models)
-\end{itemize}
-
-\textbf{Contributions}
-Comparative study of popular active learning query strategies using modAL framework. Contributions are as follows:
-\begin{itemize}
-	\item systemativ evaluation on two datasets: MNIST (Logistic Regression) and CIFAR (CNN)
-	\item analysis across different scenarios: small, moderate and large initial datasets
-	\item insight into performance impact of model complexity and initial dataset size on active learning performance
-\end{itemize}
-
-\textbf{Structure of the paper}
-\begin{itemize}
-	\item outline paper
-\end{itemize}
+\textcolor{red}{Should I even give a structure of the paper?}
 
@@
 Metrics for evaluation (Accuracy, train error, confusion matrix, ...)
-
-Tool: modAL, scikit
 

### 2025-01-09 doc/paper.tex
Finish first draft of introduction and background

@@
 It compares query strategies for a simple logistic regression model and a more complex CNN to study the effect of model complexity on the optimal query strategy.
-\textcolor{red}{CITATION were able to show that [QUERY METHODS] work well on simple models but do not generalize well on more complex models.
-To guarantee real-world applicability, knowledge about this needs to be gained as model complexities vary across different tasks and datasets}.
+\\
+It is uncertain whether query strategies which work well on simple models generalize well to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}.
+To guarantee real-world applicability, knowledge about this needs to be gained as model complexities vary across different tasks and datasets.
 Furthermore, the impact of different sizes of the initial annotated dataset is additionally investigated which is something that few studies analyze in combination.
-The effect of the combination between model complexity and initial annotated dataset size is a research gap in previously mentioned papers.
+The effect of the combination between model complexity and initial annotated dataset size is a gap not researched in previously mentioned papers.
 \\
 By examining different starting sizes, this work simulates various real-world scenarios: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
-% FROM HERE THIS IS NOT GOING TO BE PART OF THE FINAL PART
-Previous papers do not: Is everything here now mentioned previously? if yes, then it can be deleted.
-\begin{itemize}
-	\item \cite{schröder_revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
-	\item \cite{comparativesurveydeepactive}: Does not compare impact of different initial train set size and strategies on shallow classifiers.
-
-	\item According to \cite{ueno_benchmarkingofquerystrategies}: \textcolor{red}{check if true:} Does not compare different sizes of the initial starting sets as well as how simple and more complex perform with the query methods
-
-	\item \cite{werner_comparableactivelearning}: They focus on a wide range of domains, this paper only compares methods in the image domain.
-I compare strategies for a simple logistic regression model and a complex CNN across different starting sizes of the annotated dataset.
 \\
-\end{itemize}
-% TO HERE
+\\
 The contributions of this paper are as follows:
@@
 
+\newpage
+
 \section{Cool ideas}
@@
 	\item Show the typical framework of active learning in a figure in the introduction
+	\item Accuracy vs. fraction of data size in relation to the datset (percentage of used data points)
 \end{itemize}
@@
 bution. As such, it is often thought of as a measure of uncertainty or impurity in machine learning.
+
+\section{Comparison to the other papers}
+% FROM HERE THIS IS NOT GOING TO BE PART OF THE FINAL PART
+Previous papers do not: Is everything here now mentioned previously? if yes, then it can be deleted.
+\begin{itemize}
+	\item \cite{schröder_revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
+	\item \cite{comparativesurveydeepactive}: Does not compare impact of different initial train set size and strategies on shallow classifiers.
+
+	\item According to \cite{ueno_benchmarkingofquerystrategies}: \textcolor{red}{check if true:} Does not compare different sizes of the initial starting sets as well as how simple and more complex perform with the query methods
+
+	\item \cite{werner_comparableactivelearning}: They focus on a wide range of domains, this paper only compares methods in the image domain.
+I compare strategies for a simple logistic regression model and a complex CNN across different starting sizes of the annotated dataset.
+\\
+\end{itemize}
+% TO HERE
 

### 2025-01-09 doc/paper.tex
Finish Introduction and Related works

@@
 
-\section{Introduction}\label{sec:intro}
-Complex tasks in supervised learning usually require large annotated datasets.
-However, the cost and time needed to label samples can be high, especially when expert knowledge in the according field is required.
+\section{Introduction and Related Work}\label{sec:intro}
+Supervised learning often requires large, annotated datasets, but labeling data can be costly and time-intensive, especially when expert knowledge is needed. 
+Active learning addresses this challenge of reducing the overall labeling costs through the selection of the most informative samples for annotation based on a query strategy.
+This approach enables machine learning models to perform well in tasks with limited data and high labeling costs.
 \\
-Active learning aims to reduce the overall labeling cost by selecting the most informative samples for annotation by an oracle and thus enable using machine learning for tasks where annotated data is sparse.
+The effectiveness of active learning depends heavily on the chosen query strategy, which determines which samples are selected for annotation. Two main frameworks are commonly used:
+\begin{enumerate}
+	\item Pool-based active learning: A model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data.
+The most informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set.
+The model is retrained with the updated training set and the process is repeated until a stopping criterion is met.
+	\item Stream-based active learning: Here, unlabeled data arrives sequentially and the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of its prediction and checking whether a certain threshold is reached.
+If not, the instance is discarded.
+\end{enumerate}
+This report focuses on comparing several query strategies for both pool-based and stream-based active learning.
+Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
 \\
-The effectiveness of active learning relies heavily on the choice of query strategy, as it determines which samples are selected for annotation and thus directly influences model performance.
-In pool-based active learning, a model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data.
-Informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set.
-The model is retrained, and the process is repeated until a stopping criterion is met.
-\\
-In stream-based active learning, unlabeled data arrives sequentially.
-Thus, the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of the prediction and checking whether a threshold is reached.
-If not, the instance is discarded.
-\\
-This report focuses on comparing several pool-based as well as stream-based query strategies.
-The modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning} is used in the following experiments as it simplified the implementation of active learning pipelines, contains multiple already implemented querying strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
-\\
-Several previous works explore different query methods for different tasks. \\
-\cite{schröder_revisitinguncertaintybasedquerystrategies} compares various uncertainty based query strategies in the context of fine-tuning transformer models for text classification, while \cite{zhan_comparativesurveydeepactive} compares several query methods on MNIST and CIFAR.
+Several previous works have explored and compared different query methods across various domains. \\
+\cite{schröder_revisitinguncertaintybasedquerystrategies} compare various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks.
+\cite{zhan_comparativesurveydeepactive} on the other hand compares several query strategies on the image datasets MNIST and CIFAR.
 \\
 According to \cite{ueno_benchmarkingofquerystrategies} most papers concentrate on two image-based datasets: MNIST~\cite{mnist} and CIFAR~\cite{cifar}, while their paper evaluates strategies on six different datasets, including medical and visual inspection images.
-They also verify the effectiveness of query strategies independently of the training process, underfitting and randomness due to early-stage training in an experiment where fully trained models are used to select the most informative samples according to various query strategies and to construct a labeled dataset.
+They also examine strategies independent of training processes to minimize biases from early-stage randomness or underfitting.
+This is shown in an experiment where fully trained models are used to select the most informative samples based on several different query strategies to construct a labeled dataset.
 \\
-\cite{werner_comparableactivelearning} agrees with their statement that experimental settings are not standardized across papers which makes comparisons of different existing and new query strategies difficult.
-They propose a benchmark framework evaluating active learning strategies on tabular, image, and text datasets using robust evaluation protocols to mitigate variance and ensure comparability.
+A recurring issue in active learning research is the lack of standardization in experimental setups, as noticed additionally by~\cite{werner_comparableactivelearning}.
+This inconsistency complicates the comparability of existing and novel query strategies.
+To address this, they propose a benchmark framework for evaluatting active learning strategies across tabular, image and text datasets.
+This framework uses robust evaluation protocols to reduce variance and ensure comparability.
 \\
-While \cite{ueno_benchmarkingofquerystrategies} evaluates strategies across several image-based datasets from a wide range of image-domains like the medical and visual inspection images, \cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets.
+In contrast, \cite{ueno_benchmarkingofquerystrategies} focus specifically on image-based datasets from diverse domains such as medical or visual inspection images.
+While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets, this report focuses on comparing strategies in the image domain.
+It compares query strategies for both models such as simple logistic regression on MNIST and more complex architectures, like a CNN on CIFAR-10 to study the effect of model complexity on the optimal query strategy.
 \\
-This report focuses on comparing strategies in the image domain.
-It compares query strategies for a simple logistic regression model and a more complex CNN to study the effect of model complexity on the optimal query strategy.
+It is uncertain whether well working query strategies generalize well from simple to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}.
+Understanding this generalizability is crucial for real-world applications where model complexity varies across tasks and datasets.
 \\
-It is uncertain whether query strategies which work well on simple models generalize well to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}.
-To guarantee real-world applicability, knowledge about this needs to be gained as model complexities vary across different tasks and datasets.
-Furthermore, the impact of different sizes of the initial annotated dataset is additionally investigated which is something that few studies analyze in combination.
-The effect of the combination between model complexity and initial annotated dataset size is a gap not researched in previously mentioned papers.
-\\
-By examining different starting sizes, this work simulates various real-world scenarios: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
-\\
+Furthermore, this report investigates the impact of the initial labeled dataset size, as its impact in combination with model complexity is rarely explored. 
+By examining scenarios with varying amounts of initial labeled data, this work simulates various real-world applications: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
 \\
@@
 \begin{itemize}
-	\item \textbf{Systematic evaluation of active learning query strategies} on MNIST with a logistic regression model and CIFAR-10 using a CNN to analyze their behavior across datasets and model complexities.
-	\item \textbf{Investigation of model complexity's influence} on the effectiveness of query strategies to provide knowledge into how shallow and deep models respond differently to active learning methods.
-	\item \textbf{Analysis of the initial labeled dataset size's impact} on the performance of query strategies to simulate real-world scenarios such as starting completely from scratch, having a small labeled dataset or fine-tuning a model.
+	\item \textbf{Systematic evaluation of active learning query strategies} across different datasets and model complexities, using MNIST with logistic regression and CIFAR-10 with a CNN.
+	\item \textbf{Investigation of the incluence of model complexity} on the effectiveness of query strategies to provide insights into how shallow and deep models respond differently to active learning methods.
+	\item \textbf{Analysis of impact of initial labeled dataset size} on the performance of query strategies, simulating possible real-world scenarios such as starting from scratch, having a small labeled dataset or fine-tuning a pre-trained model.
 \end{itemize}
-
-\textcolor{red}{Should I even give a structure of the paper?}
+Section~\ref{sec:methods} outlines the used datasets, processing steps and models.
+Section~\ref{sec:experiments} describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively.
+The report concludes with key findings in section~\ref{sec:conclusion}
 

### 2025-01-10 doc/paper.tex
Work on report and add cnn model and training pipeline
This was pain

@@
 Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
-\\
+\textcolor{red}{CNN implemented with PyTorch(ADD CITATION) and transformed with skorch~\cite{skorch} to make it compatible with modAL as it is only compatible with sklearn models.} \\
 Several previous works have explored and compared different query methods across various domains. \\
@@
 \\
+\\
 The contributions of this paper are as follows:
@@
 \section{Methods}\label{sec:methods}
-\label{data_and_methods}
+Several different datasets are used in the following experiments.
 \subsection{Datasets}
-\begin{itemize}
-	\item MNIST and CIFAR describe
-	\item processing steps \\
-	MNIST: Vectorized, flattened (describe input size), normalized
-\end{itemize}
+MNIST is a dataset consisting of handwritten digits with 60.000 train samples and 10.000 test samples.
+Digits are size-normalized and centered in a $28 \times 28$ image~\cite{mnist}.
+Each digits belongs to one of 10 classes, corresponding from 0 to 9.
+\\
+Upon loading the dataset, each image is vectorized, i.e.
+transformed from a $28 \times 28$ sized image-matrix to a vector of length $784$.
+Furthermore, each pixel contains a grayscale-value from 0 to 255.
+Pixel values are normalized to a range between 0 and 1 to provide a value range that works well with logistic regression.
+\\
+\textcolor{red}{Here describe the validation split}
+\\
+\\
+CIFAR-10 is a dataset which consists of 50.000 train- and 10.000 test-images.
+Each natural image belongs to one of 10 classes.
+Each class has exactly 5.000 train and 1.000 test images.
+The image-size is $32 \times 32$ pixels.
+\textcolor{red}{CIFAR preprocessing split and validation split}
+\\
+\\
+Both used datasets are balanced, i.e. each class has roughly the same number of samples as any other class.
 
@@
 	\item Logistic Regression: Details on implementation and hyperparameters, how were they chosen?
-	\item same for CNN
+	\item same for CNN: CrossEntropyLoss (combines Log Softmax and NLLoss in one single class)
 \end{itemize}
+Are models re-initialized after each new data points and newly trained or fine-tuned?
 
@@
 \\
-Most methods require knowledge about the balance of the dataset and might be disadvantageous to use in case of a disbalanced dataset. In real world scenarios, this is not known. \\
+Most methods require knowledge about the balance of the dataset and might be disadvantageous to use in case of a disbalanced dataset.
+In real world scenarios, this is not known.
+Both used datasets are balanced so the impact of query strategies on imbalanced datasets was not tested.
+\\
 Only image-based datasets were tested. Not vector- or text-based.

### 2025-01-10 code/cnn_model.py

@@ -1,0 +1,50 @@
+import math
+from skorch import NeuralNetClassifier
+import torch
+from torch import nn
+
+
+class CNN(nn.Module):
+    def __init__(self, model_params):
+        super(CNN, self).__init__()
+        self.model_params = model_params
+        self.conv1 = nn.Sequential(
+            nn.Conv2d(model_params['num_channels'], model_params['l1_channels'],
+                      kernel_size=model_params['l1_kernel_size']),
+            nn.ReLU()
+        )
+        self.conv2 = nn.Sequential(
+            nn.Conv2d(model_params['l1_channels'], model_params['l2_channels'],
+                      kernel_size=model_params['l2_kernel_size']),
+            nn.ReLU(),
+            nn.MaxPool2d(kernel_size=model_params['l2_max_pool_kernel_size']),
+            nn.Dropout(model_params['l2_dropout'])
+        )
+
+        self.l3_input = self.calculate_l3_input(model_params['input_size'])
+        self.fcs = nn.Sequential(
+            nn.Linear(in_features=self.l3_input, out_features=model_params['l4_input']),
+            # Adjust dimensions based on input size
+            nn.ReLU(),
+            nn.Dropout(model_params['l3_dropout']),
+            nn.Linear(in_features=model_params['l4_input'],
+                      out_features=model_params['l5_input']),
+            nn.ReLU(),
+            nn.Dropout(model_params['l4_dropout']),
+            nn.Linear(in_features=model_params['l5_input'],
+                      out_features=model_params['output_size'])
+        )
+
+    def calculate_l3_input(self, input_size):
+        dummy_input = torch.zeros(1, self.model_params['num_channels'], input_size, input_size)
+        x = self.conv1(dummy_input)
+        x = self.conv2(x)
+        return int(math.prod(
+            x.shape[1:]))  # product of all the elements in the flattened vector except for the batch size of course
+
+    def forward(self, x):
+        x = self.conv1(x)
+        x = self.conv2(x)
+        x = x.view(x.size(0), -1)  # Flatten the output
+        x = self.fcs(x)
+        return x

### 2025-01-10 code/utils.py

@@ -3,11 +3,22 @@ from sklearn.datasets import fetch_openml
 from sklearn.metrics import accuracy_score, log_loss
 from sklearn.linear_model import LogisticRegression
 import numpy as np
+import tensorflow as tf
+import torch
+from skorch import NeuralNetClassifier
+from torch import nn
+from cnn_model import CNN
 
 
 def initialize_random_number_generators(seed):
     np.random.seed(seed)
     random.seed(seed)
+    # Cuda should not be used but I found it hard to produce reproducible results with
+    # Pytorch models so I'll set this as a precaution
+    torch.manual_seed(seed)
+    torch.cuda.manual_seed_all(seed)  # For CUDA
+    torch.backends.cudnn.deterministic = True  # Ensure deterministic behavior
+    torch.backends.cudnn.benchmark = False  # Ensure reproducibility
 
 
 def load_MNIST():
@@ -16,8 +27,8 @@ def load_MNIST():
     X, y = mnist.data.to_numpy(), mnist.target.to_numpy()
 
     # Normalize as data ranges between 0 and 255
-    X = X / 255
-    # Convert type of y from string to integer
+    X = X / 255.0
+    # Convert type of y (aka the label) from string to integer
     y = y.astype(int)
 
     X_train, y_train = X[:60000], y[:60000]
@@ -26,10 +37,56 @@ def load_MNIST():
     return X_train, y_train, X_test, y_test, X, y
 
 
-def log_metrics(step: int, model, X_train, y_train, X_test, y_test, metrics: dict):
-    train_loss = log_loss(y_train, model.predict_proba(X_train))
-    test_loss = log_loss(y_test, model.predict_proba(X_test))
-    test_accuracy = accuracy_score(y_test, model.predict(X_test))
+def load_CIFAR():
+    (X_train, y_train), (X_test, y_test) = tf.keras.datasets.cifar10.load_data()
+
+    # Normalize pixel values as they are rgb values between 0 and 255
+    X_train = X_train.astype('float32') / 255.0
+    X_test = X_test.astype('float32') / 255.0
+
+    # Transpose the datasets to have the channel (3) first and then height (1) and width (2)
+    X_train = np.transpose(X_train, (0, 3, 1, 2))
+    X_test = np.transpose(X_test, (0, 3, 1, 2))
+
+    # Convert labels to 1D array
+    y_train, y_test = y_train.flatten(), y_test.flatten()
+
[…]

### 2025-01-12 doc/paper.tex
Todays progress: Add script for cross validation and add n_itial as a percentage of the train set

@@
 Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
-\textcolor{red}{CNN implemented with PyTorch(ADD CITATION) and transformed with skorch~\cite{skorch} to make it compatible with modAL as it is only compatible with sklearn models.} \\
+\textcolor{red}{Convolutional Neural Netowrk (CNN) implemented with PyTorch(ADD CITATION) and transformed with skorch~\cite{skorch} to make it compatible with modAL as it is only compatible with sklearn models.} \\
 Several previous works have explored and compared different query methods across various domains. \\
@@
 \\
-\textcolor{red}{Here describe the validation split}
-\\
 \\
@@
 The image-size is $32 \times 32$ pixels.
-\textcolor{red}{CIFAR preprocessing split and validation split}
+\textcolor{red}{CIFAR preprocessing split}. CIFAR-10 includes real-world images and is thus expected to be more complicated to learn.
+
+Both used datasets are balanced, i.e.
+each class has roughly the same number of samples as any other class.
+Furthermore, 20\% of both train sets are used for 5-fold-cross-validation, resulting in a test set size of $48.000$ train images for MNIST and $40.000$ train images for CIFAR.
+
+\subsection{Multinomial Logistic Regression Model}
+Sklearn's LogisticRegression module is used to predict digits on MNIST.
+Multinomial logistic regression predicts a probability for each class instead of a single probability as in Binomial logistic regression.
 \\
+The model estimates the probability $P(y=i \mid X)$ of a sample belonging to class $i$ using the SoftMax function:
+\begin{align}
+	P(y=i \mid X)=\frac{\exp \left(X \beta_i\right)}{\sum_{j=1}^k \exp \left(X \beta_j\right)}
+\end{align}
+where $\beta_i$ are the coefficients for class $i$. The SoftMax function ensures that probabilities across classes sum to 1.
+
+The model maximizes the likelihood of the observed data:
+\begin{align}
+	\mathcal{L}(\beta)=\prod_{n=1}^N \prod_{i=1}^k\left(P\left(y_n=i \mid X_n\right)\right)^{y_{n i}}
+\end{align}
+where $N$ is the number of samples, and $y_{n i}$ indicates if sample $n$ belongs to class $i$. \textcolor{red}{Should I also explain Cross-Entropy loss? $\begin{aligned}
+		&\text { For a multinomial logistic regression model with } k \text { classes, the cross-entropy loss } \mathcal{L} \text { is defined as: }\\
+		&\mathcal{L}=-\frac{1}{N} \sum_{n=1}^N \sum_{j=1}^k y_{n j} \log \left(P\left(y=j \mid X_n\right)\right)
+\end{aligned}$}\\
+Regularization using the $\ell_2$-norm prevents overfitting with a strength $C=0.01$.
+The lbfgs-solver, with a maximum iteration of 500, is used for efficient gradient-based optimization.
+Early-stopping with a tolerance of 0.001 is used.
 \\
-Both used datasets are balanced, i.e. each class has roughly the same number of samples as any other class.
 
-\subsection{Models}
-\begin{itemize}
-	\item Logistic Regression: Details on implementation and hyperparameters, how were they chosen?
-	\item same for CNN: CrossEntropyLoss (combines Log Softmax and NLLoss in one single class)
-\end{itemize}
-Are models re-initialized after each new data points and newly trained or fine-tuned?
+\subsection{Convolutional Neural Network}
+The model which is used to learn the classes in CIFAR-10 is a CNN. It comprises of two convolutional layers followed by three fully connected layers. \\
+The first convolutional layer processes the input image with 3 input and $l1\_channels$ output channels, followed by a ReLU activation.
+The second convolutional layer further processes the output from the first layer with $l1\_channels$ input and $l2\_channels$ output channels, applying ReLU activation, max pooling, and dropout for regularization.
+\\
+The output from the convolutional layers is flattened and passed through three fully connected layers.
+The first two fully connected layers are followed by ReLU activations and dropout for regularization.
+The first fully connected layer has an output size of $l4_input$ and the second layer has an output size of $l5_input$ neurons.
+The final output layer has 10 neurons, corresponding to the number of classes in the CIFAR dataset.
+\\
+The model is implemented using PyTorch\textcolor{red}{cite} and skorch~\cite{skorch} is used to make it possible to use it with Sklearn, allowing integration with the modAL active learning framework.
+\\
+\textcolor{red}{Early stopping?}
+\\
+During training, PyTorch's CrossEntropyLoss is used.
 
@@
 \begin{itemize}
-	\item How are the splits (initial pool, initial training data) generated
-	\item numbmer of iterations for active learning
-	\item Used active learning query strategies
+	\item During active learning with pool-based query strategies in the case of logistic regression, 5 instances are queried at a time for 250 iterations resulting in a labeled pool which is increased by $1.250$ samples.
+\\
+	When training the CNN, 250 instances are chosen for 50 iterations. This enlarges the labeled pool by $12.500$ samples. \\
+	In stream-based methods, the number of queries is adapted to match the given numbers and guarantee that the labeled dataset size is the same as in the pool based methods. 
+	\item Used active learning query strategies include pool- and stream-based methods.
+The pool based methods include uncertainty sampling, uncertainty margin sampling and uncertainty entropy sampling.
+Stream-based active learning is done with uncertainty as the evaluation metric deciding whether to use samples or not.
+Additionally, random sampling was used to serve as a reference point.
+	\textcolor{red}{maybe explain the query strategies more in detail}
+	\item After querying instances and retraining, models are not re-initialized but rather fine-tuned. 
 \end{itemize}
 
-Metrics for evaluation (Accuracy, train error, confusion matrix, ...)
+\section{Experiments}\label{sec:experiments}
+Several experiments are conducted with the two model-dataset combinations to research different scenarios for which active learning can be used.
+\begin{itemize}
+	\item Experiment 1: Starts with a minimal amount of data. In both cases these are 10 samples.
+	\item Experiment 2: Starts with $15\%$ of the train dataset. 
+	\item Experiment 3: Starts with $80\%$ of the train dataset.
+\end{itemize}
+In experiment 1, a random sample is chosen for each class in the datasets. \\
+In each experiment, the random samples
 
-\section{Experiments}\label{sec:experiments}
-Setup: details for reproducibility (random number generators)?
-Random seed = 42
-\cite{comparableactivelearning} mention reproducibility problems in models even when setting random number generators before each run.
-By training models multiple times in succession and checking whether the same configurations lead to different outcomes, it is shown that in this scenario, reproducibility is guaranteed.
-\textcolor{red}{TODO: Schnell ein skript aufsetzen und dann finale accuracy und so weiter vergleichen.}
+For reproducibility purposes, the same random seed $42$ was used to initialize all random number generators and model weights to ensure that same configurations lead to the same outcomes, even when training models in succession.
+
+For each query method one model is trained. (I did not perform multiple runs)
+Before starting the active learning process, each model is pre-trained using the samples in the initial labeled dataset for \textcolor{red}{X epochs}.
+In the case of pool-based query methods, 5 samples are queried for 250 iterations when training with samples from MNIST and 250 samples are queried for 50 iterations when training with instances from CIFAR.
 \\
-Did I perform multiple runs?
-\\
+\begin{itemize}
+	\item Batch size for CNN is 128.
+	\item Evaluation metrics. Metrics for evaluation (Accuracy, train error, confusion matrix, ...)
+	\item How are the splits (initial pool, initial training data) generated
+	\item initial labeled pool sizes
+\end{itemize}
 
@@
 Only image-based datasets were tested. Not vector- or text-based.
+
+Only one network is trained for each query method. This could be improved upon when training multiple models.
 

### 2025-01-12 code/utils.py

@@ -5,6 +5,7 @@ from sklearn.linear_model import LogisticRegression
 import numpy as np
 import tensorflow as tf
 import torch
+from sklearn.model_selection import train_test_split
 from skorch import NeuralNetClassifier
 from torch import nn
 from cnn_model import CNN
@@ -21,7 +22,7 @@ def initialize_random_number_generators(seed):
     torch.backends.cudnn.benchmark = False  # Ensure reproducibility
 
 
-def load_MNIST():
+def load_MNIST(random_seed, validation_split=0.2):
     mnist = fetch_openml('mnist_784', version=1, parser="auto")
 
     X, y = mnist.data.to_numpy(), mnist.target.to_numpy()
@@ -31,10 +32,15 @@ def load_MNIST():
     # Convert type of y (aka the label) from string to integer
     y = y.astype(int)
 
-    X_train, y_train = X[:60000], y[:60000]
+    X_train_full, y_train_full = X[:60000], y[:60000] # This is the default way of extracting train and test data
     X_test, y_test = X[60000:], y[60000:]
 
-    return X_train, y_train, X_test, y_test, X, y
+    # Create validation split
+    X_train, X_val, y_train, y_val = train_test_split(
+        X_train_full, y_train_full, test_size=validation_split, random_state=random_seed
+    )
+
+    return X_train, y_train, X_test, y_test, X_val, y_val, X, y
 
 
 def load_CIFAR():
@@ -97,7 +103,7 @@ def log_metrics(step: int, model, X_train, y_train, X_test, y_test, metrics: dic
     return
 
 
-def create_log_reg_model(model_params, random_seed):
+def create_log_reg_model(model_params, random_seed, device="cpu"):
     return LogisticRegression(solver='sag',
                               penalty=model_params["regularization"],
                               C=model_params["regularization_strength"],
@@ -110,11 +116,6 @@ def create_log_reg_model(model_params, random_seed):
 def create_cnn_model(model_params, random_seed, device="cpu"):
     initialize_random_number_generators(random_seed)
     cnn = CNN(model_params)
-    optimizer = torch.optim.Adam(
-        cnn.parameters(),
-        lr=model_params["learning_rate"],
-        weight_decay=model_params["weight_decay"]
-    )
     return NeuralNetClassifier(cnn,
                criterion=nn.CrossEntropyLoss, # don't i need to use another loss?
                optimizer=torch.optim.Adam,  # Pass the optimizer class, not an instance
[…]

### 2025-01-13 code/utils.py
Finish parameter search and add parameters to model and file

@@ -104,11 +104,12 @@ def log_metrics(step: int, model, X_train, y_train, X_test, y_test, metrics: dic
 
 
 def create_log_reg_model(model_params, random_seed, device="cpu"):
-    return LogisticRegression(solver='sag',
+    return LogisticRegression(solver=model_params['solver'],
                               penalty=model_params["regularization"],
                               C=model_params["regularization_strength"],
                               multi_class='multinomial',
                               max_iter=model_params["max_iterations_per_epoch"],
+                              tol=model_params["early_stopping_tol"],
                               warm_start=True,
                               random_state=random_seed)
 

### 2025-01-13 doc/paper.tex
refactor, update numbers in paper

@@
 	\item Experiment 1: Starts with a minimal amount of data. In both cases these are 10 samples.
-	\item Experiment 2: Starts with $15\%$ of the train dataset. 
-	\item Experiment 3: Starts with $80\%$ of the train dataset.
+	\item Experiment 2: Starts with $10\%$ of the train dataset, resulting in $4800$ initial samples for MNIST and $4000$ \textcolor{red}{überprüfen} for CIFAR.
+	\item Experiment 3: Starts with $50\%$ of the train dataset. Thus, $24.000$ labeled instances are chosen for MNIST and $20.000$ for CIFAR. \textcolor{red}{überprüfen ob das sinn ergibt}
 \end{itemize}
@@
 	\item initial labeled pool sizes
+	\item stream based: if 10.000 instances are pooled but none are accepted, training is finished early.
 \end{itemize}

### 2025-01-13 code/utils.py

@@ -1,4 +1,8 @@
 import os, random, pickle
+import time
+
+from modAL import ActiveLearner
+from modAL.uncertainty import classifier_margin
 from sklearn.datasets import fetch_openml
 from sklearn.metrics import accuracy_score, log_loss
 from sklearn.linear_model import LogisticRegression
@@ -158,3 +162,162 @@ def load_file(path: str):
         loaded_dictionary = pickle.load(file)
 
     return loaded_dictionary
+
+
+# Training scripts:
+""" 
+    Method for pool-based active learning query strategies. 
+"""
+
+
+def train_active_learner(model_params, query_strat, n_query_instances: int, epochs: int, random_seed: int,
+                         datasets, create_model, device="cpu"):
+    initialize_random_number_generators(seed=random_seed)
+
+    dataset_name = datasets['dataset_name']
+    X_initial, y_initial = datasets['X_initial'], datasets['y_initial']
+    X_train, y_train = datasets['X_train'], datasets['y_train']
+    X_test, y_test = datasets['X_test'], datasets['y_test']
+    pool_idx = datasets['pool_idx']
+    X_pool, y_pool = X_train[pool_idx], y_train[pool_idx]
+
+    is_cnn = (dataset_name == "CIFAR")
+
+    model = create_model(model_params, random_seed=random_seed, device=device)
+
+    learner = ActiveLearner(estimator=model,
+                            query_strategy=query_strat,
+                            X_training=X_initial,
+                            y_training=y_initial)
+
+    # Passing X_training and y_training to the ActiveLearner automatically calls the fit method for log_reg with these data points
+
+    metrics = {'queries': [], 'train_loss': [], 'train_loss_current': [], 'test_loss': [], 'test_acc': []}
+
+    start = time.time()
+    for epoch in range(epochs):  # epochs=n_queries to have both models trained on the same number of overall epochs
+        query_idx, query_inst = learner.query(X_pool, n_instances=n_query_instances)
+
+        # Simulate labeling
+        learner.teach(X_pool[query_idx], y_pool[query_idx], only_new=False)
+
+        # Remove queried point(s)
+        X_pool, y_pool = np.delete(X_pool, query_idx, axis=0), np.delete(y_pool, query_idx, axis=0)
+
+        metrics['queries'].append(query_idx)
+        # log_metrics logs train loss for the whole train dataset which doesn't reflect the actual value in the current step but gives the ability to compare both models on the training set.
[…]

### 2025-02-03 doc/paper.tex
Add finished report

--- 2025-01-10 doc/paper.tex
@@
 \twocolumn[
-\icmltitle{TODO: Funky active learning pun}
+\icmltitle{Active Learning: Because Guesswork is Overrated}
 
@@
 \begin{icmlauthorlist}
-	\icmlauthor{Lukas Weber}{first}
+	\icmlauthor{Lukas Weber, lukas.weber@studenti.unitn.it}{first}
 \end{icmlauthorlist}
@@
 %\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 
 \begin{abstract}
-	A lot of unlabelled data generally available but annotation is costly.
-A goal in fields where labelled data is not available  is to maximize performance with minimum amount of labeled data instances.
-\\
-	Active learning selects instances for labelling based on a selection strategy. 
-	\begin{itemize}
-		\item Objective: Compare active learning query methods using modAL.
-		\item Methods: Experiments with Logistic Regression (MNIST) and a CNN (CIFAR).
-		\item Experiments: Vary initial dataset sizes (small, moderate, large).
-		\item Key findings: Highlight main conclusions about performance in different scenarios.
-	\end{itemize}
+	A huge amount of unlabeled data is typically available but annotating large datasets is often costly and time-consuming, making active learning a crucial tool for selecting the most informative samples to minimize annotation efforts.
+This report compares different active learning query strategies using the modAL framework in a pool-based setting.
+In experiments, a logistic regression model and an ensemble of such models is evaluated on the MNIST dataset.
+The impact of different query strategies is analyzed, including Uncertainty Sampling, Margin Sampling, Entropy Sampling (including ensemble adaptions) and Ranked batch-mode Sampling.
+Findings indicate that querying samples with Margin Sampling achieves the highest accuracy and fastest convergence.
+While ensemble models show faster convergence with small datasets, they provide shrinking improvements over single models as dataset size increases.
+These results highlight the high importance of query strategy selection on model performance in active learning pipelines and explore conditions under which ensemble models may improve performance.
 \end{abstract}
@@
 \section{Introduction and Related Work}\label{sec:intro}
-Supervised learning often requires large, annotated datasets, but labeling data can be costly and time-intensive, especially when expert knowledge is needed. 
-Active learning addresses this challenge of reducing the overall labeling costs through the selection of the most informative samples for annotation based on a query strategy.
+Supervised learning often requires large, annotated datasets.
+However obtaining high-quality labeled data can be costly and time-intensive, especially expert knowledge is required, such as medical imaging.
+Though, unlabeled data is usually easily available.
+Active learning addresses this challenge by reducing the overall labeling costs through the selection of most informative samples for annotation based on a query strategy.
 This approach enables machine learning models to perform well in tasks with limited data and high labeling costs.
 \\
-The effectiveness of active learning depends heavily on the chosen query strategy, which determines which samples are selected for annotation. Two main frameworks are commonly used:
-\begin{enumerate}
-	\item Pool-based active learning: A model is trained on a small labeled dataset and used to evaluate a large pool of unlabeled data.
-The most informative samples are selected based on a query strategy, labeled by an oracle, and added to the training set.
+The effectiveness of active learning is tied to the chosen query strategy, that determines which unlabeled samples are selected for annotation and then added to the training set. \\
+This report evaluates several strategies in a pool-based active learning setting: A model is initially trained on a small labeled dataset and used to evaluate a large pool of unlabeled data.
+The most informative samples are selected based on a query strategy, they are then labeled by an oracle, and added to the training set.
 The model is retrained with the updated training set and the process is repeated until a stopping criterion is met.
-	\item Stream-based active learning: Here, unlabeled data arrives sequentially and the model decides whether to query the label of an incoming instance using an evaluation metric like the uncertainty of its prediction and checking whether a certain threshold is reached.
-If not, the instance is discarded.
-\end{enumerate}
-This report focuses on comparing several query strategies for both pool-based and stream-based active learning.
-Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplified the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with models from the Scikit-learn library~\cite{scikit-learn}.
-\textcolor{red}{CNN implemented with PyTorch(ADD CITATION) and transformed with skorch~\cite{skorch} to make it compatible with modAL as it is only compatible with sklearn models.} \\
-Several previous works have explored and compared different query methods across various domains. \\
-\cite{schröder_revisitinguncertaintybasedquerystrategies} compare various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks.
-\cite{zhan_comparativesurveydeepactive} on the other hand compares several query strategies on the image datasets MNIST and CIFAR.
+Experiments in this paper leverage the modular Active Learning framework (modAL)~\cite{danka_modalmodularactivelearning}, which simplifies the implementation of active learning pipelines, includes multiple pre-implemented query strategies and is compatible with Scikit-learn models~\cite{scikit-learn}.
+\\
+The experiments focus on the image domain, using the MNIST dataset~\cite{mnist}, and examine the impact of query strategies on a single logistic regression model and a committee-based approach.
+Several previous works have explored and compared different query methods across various domains.
+\cite{schröder_revisitinguncertaintybasedquerystrategies} compared various uncertainty-based query strategies in the context of fine-tuning transformer models in text classification tasks, while \cite{zhan_comparativesurveydeepactive} examined several query methods in image classification tasks using MNIST and CIFAR.
 \\
@@
 \\
-A recurring issue in active learning research is the lack of standardization in experimental setups, as noticed additionally by~\cite{werner_comparableactivelearning}.
+A recurring issue in active learning research is the lack of standardization in experimental setups, making it difficult to compare existing and new query strategies, as noticed additionally by~\cite{werner_comparableactivelearning}.
 This inconsistency complicates the comparability of existing and novel query strategies.
-To address this, they propose a benchmark framework for evaluatting active learning strategies across tabular, image and text datasets.
+To address this, they propose a benchmark framework for evaluating active learning strategies across tabular, image and text datasets.
 This framework uses robust evaluation protocols to reduce variance and ensure comparability.
@@
 In contrast, \cite{ueno_benchmarkingofquerystrategies} focus specifically on image-based datasets from diverse domains such as medical or visual inspection images.
-While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets, this report focuses on comparing strategies in the image domain.
-It compares query strategies for both models such as simple logistic regression on MNIST and more complex architectures, like a CNN on CIFAR-10 to study the effect of model complexity on the optimal query strategy.
-\\
-It is uncertain whether well working query strategies generalize well from simple to more complex models like Neural Networks~\cite{schröder_surveyactivelearningtext}.
-Understanding this generalizability is crucial for real-world applications where model complexity varies across tasks and datasets.
-\\
-Furthermore, this report investigates the impact of the initial labeled dataset size, as its impact in combination with model complexity is rarely explored. 
-By examining scenarios with varying amounts of initial labeled data, this work simulates various real-world applications: (1) starting with no labeled data, (2) having a small pool of labeled data, and (3) working with a well-trained model to identify data points for fine-tuning.
+While~\cite{werner_comparableactivelearning} focuses on a wide range of data-domains including images, text and vector-based datasets.
+This work focuses on query strategy comparisons within the image domain, specifically for a logistic regression model and an ensemble-based approach on MNIST. \\
+The contributions of this paper are as follows:
+\begin{itemize}
+	\item A systematic evaluation of active learning query strategies using \textbf{logistic regression on MNIST}.
+	\item \textbf{Investigation of the incluence of model aggregation} on the effectiveness of query strategies to provide insights into how these models respond differently to active learning methods.
+\end{itemize}
+Section~\ref{sec:methods} outlines the used dataset, processing steps and models and describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively.
+The report concludes with key findings in section~\ref{sec:conclusion}.
 \\
 \\
-The contributions of this paper are as follows:
-\begin{itemize}
-	\item \textbf{Systematic evaluation of active learning query strategies} across different datasets and model complexities, using MNIST with logistic regression and CIFAR-10 with a CNN.
-	\item \textbf{Investigation of the incluence of model complexity} on the effectiveness of query strategies to provide insights into how shallow and deep models respond differently to active learning methods.
-	\item \textbf{Analysis of impact of initial labeled dataset size} on the performance of query strategies, simulating possible real-world scenarios such as starting from scratch, having a small labeled dataset or fine-tuning a pre-trained model.
-\end{itemize}
-Section~\ref{sec:methods} outlines the used datasets, processing steps and models.
-Section~\ref{sec:experiments} describes the experimental design, followed by the results and discussion in sections~\ref{sec:results} and~\ref{sec:discussion} respectively.
-The report concludes with key findings in section~\ref{sec:conclusion}
+The corresponding implementation can be found on \textcolor{red}{\href{https://github.com/LukaslWeber/NLP_II_Interactive_Learning}{Github}}.
 
 \section{Methods}\label{sec:methods}
-Several different datasets are used in the following experiments.
-\subsection{Datasets}
+The following section describes the dataset which was used in experiments, preprocessing steps, the logistic regression model and other parameters which are needed for reproducibility.
+\subsection{Dataset}
 MNIST is a dataset consisting of handwritten digits with 60.000 train samples and 10.000 test samples.
 Digits are size-normalized and centered in a $28 \times 28$ image~\cite{mnist}.
-Each digits belongs to one of 10 classes, corresponding from 0 to 9.
+Each digit belongs to one of 10 classes, corresponding to the  numbers from 0 to 9.
 \\
@@
 transformed from a $28 \times 28$ sized image-matrix to a vector of length $784$.
-Furthermore, each pixel contains a grayscale-value from 0 to 255.
-Pixel values are normalized to a range between 0 and 1 to provide a value range that works well with logistic regression.
-\\
-\textcolor{red}{Here describe the validation split}
-\\
-\\
-CIFAR-10 is a dataset which consists of 50.000 train- and 10.000 test-images.
+Furthermore, each pixel initially contains a grayscale value ranging from 0 to 255.
+Pixel values are additionally normalized to a range between 0 and 1 in order to provide a value range that works well with logistic regression.
+%\\
+%\\
+%CIFAR-10 is a dataset which consists of 50.000 train- and 10.000 test-images.
 Each natural image belongs to one of 10 classes.
@@
 The image-size is $32 \times 32$ pixels.
-\textcolor{red}{CIFAR preprocessing split and validation split}
+%\textcolor{red}{CIFAR preprocessing split}. CIFAR-10 includes real-world images and is thus expected to be more complicated to learn.
+%
+MNIST is a class-balanced dataset, i.e. each class has roughly the same number of samples as any other class.
+
+\subsection{Multinomial Logistic Regression Model}
+Sklearn's LogisticRegression module is used to predict digits with the given data from MNIST.
+Multinomial logistic regression predicts a probability for each possible class instead of a single probability as in Binomial logistic regression.
 \\
+The model estimates the probability $P(y=i \mid X)$ of a sample belonging to class $i$ using the SoftMax function:
+\begin{align}
+	P(y=i \mid X)=\frac{\exp \left(X \beta_i\right)}{\sum_{j=1}^k \exp \left(X \beta_j\right)}
+\end{align}
+where $\beta_i$ are the coefficients for class $i$.
+The SoftMax function ensures that probabilities across classes sum to 1 and consequently that a model creates a probability distribution over the classes.
 \\
-Both used datasets are balanced, i.e. each class has roughly the same number of samples as any other class.
-
-\subsection{Models}
-\begin{itemize}
-	\item Logistic Regression: Details on implementation and hyperparameters, how were they chosen?
-	\item same for CNN: CrossEntropyLoss (combines Log Softmax and NLLoss in one single class)
-\end{itemize}
-Are models re-initialized after each new data points and newly trained or fine-tuned?
+During training, the model maximizes the likelihood of the observed data:
+\begin{align}
+	\mathcal{L}(\beta)=\prod_{n=1}^N \prod_{i=1}^k\left(P\left(y_n=i \mid X_n\right)\right)^{y_{n i}}
+\end{align}
+where $N$ is the number of samples, and $y_{n i}$ indicates if sample $n$ belongs to class $i$. $k=10$ denotes the number of classes. \\
+The likelihood of the observed data is maximized by minimizing the Cross-Entropy loss $\mathcal{L}$ which is defined as:
+\begin{align}
+	\mathcal{L}=-\frac{1}{N} \sum_{n=1}^N \sum_{j=1}^k y_{n j} \log \left[P\left(y=j \mid X_n\right)\right]
+\end{align}
+To find hyperparameters which maximize performance on the given dataset, 5-fold-cross-validation is used with a grid-search, testing out a variety solvers, regularization methods and corresponding strengths, maximum iterations and stopping-tolerance options.
+\\
+The resulting models use the "newton-cg" solver which combines Newton's method with conjugate gradient descent.
+It uses the gradient and the second order derivative of $\mathcal{L}$ given by the Hessian matrix to iteratively find the minimum of $\mathcal{L}$.
+\\
+Regularization using the $\ell_2$-norm prevents overfitting with a strength $C=0.01$ in addition to a stopping-criterion $tol=0.0001$, which stops the learning process when the $\ell_2$-norm of the gradient of $\mathcal{L}$ is lower than $tol$, indicating convergence.
+\\
+For reproducibility purposes, the same random seed ($42$) is used to initialize all random number generators and single model weights.
+Weights of models in an ensemble are initialized with a fixed set of random seeds (42, 43, 44, 45, 46) to introduce diversity.
+These measures ensure that same configurations lead to the same outcomes, even when training models in succession.
 
 \subsection{Active Learning}
-\begin{itemize}
-	\item How are the splits (initial pool, initial training data) generated
-	\item numbmer of iterations for active learning
-	\item Used active learning query strategies
-\end{itemize}
-
-Metrics for evaluation (Accuracy, train error, confusion matrix, ...)
-
-\section{Experiments}\label{sec:experiments}
-Setup: details for reproducibility (random number generators)?
-Random seed = 42
-\cite{comparableactivelearning} mention reproducibility problems in models even when setting random number generators before each run.
-By training models multiple times in succession and checking whether the same configurations lead to different outcomes, it is shown that in this scenario, reproducibility is guaranteed.
-\textcolor{red}{TODO: Schnell ein skript aufsetzen und dann finale accuracy und so weiter vergleichen.}
+In the following experiments, a single logistic regression model is compared to a committee of 5 models across different query strategies.
+The committee predicts by averaging the class probabilities obtained by each learner, called the consensus probabilities.
+The final prediction is the class with the highest consensus probability.
 \\
-Did I perform multiple runs?
+In pool-based active learning, an initial labeled dataset is required.
+When a single model is used for training and querying new samples, $n_{init}=75$ labeled instances are used to train the model initially, ensuring a moderate fit before querying samples.
+This is done to make sure that initial samples are chosen on a meaningful basis and do not bias the rest of the training negatively.
+For the committee, each model is initialized with $60\%$ randomly chosen samples from the initial labeled train set, resulting in $45$ initial data points per model.
 \\
-
-Scenarios
-\begin{itemize}
-	\item describe experiments and their settings and purposes
-\end{itemize}
-
+For both methods, models are trained for $init\_max\_iter=50$ epochs on the initial dataset.
+$init\_max\_iter$ is based on the hyperparameters found by 5-fold-cross-validation and ensures a sufficient fit to the datapoints.
+Thus, reducing the bias from early queried samples which were chosen when model predictions are inaccurate.
+\\
+After initial training, 5 samples are queried at a time for 1000 iterations, resulting in 5000 additional samples at the end of the learning process.
+After quering, the new samples are added to the training set and models are fine-tuned for $n_{iter}=20$ iterations.
+\subsection{Query Strategies}
+A multitude of query strategies are used to compare a single model to an ensemble of models. As a control metric, Random Sampling is used to give a baseline for other query strategies.\\
+Uncertainty Sampling selects the least certain instances for labeling.
+\begin{align}
+	Uncertainty(\mathbf{x})=1-P_\theta(\hat{y} \mid \mathbf{x})
+\end{align}
+Where $\hat{y}=\operatorname{arg}\underset{y}{\operatorname{max}} \ p_\theta(y \mid \mathbf{x})$, i.e.
+corresponds to the predicted class of the model and $P_\theta$ is the probability to predict class $\hat{y}$ with the model with parameters $\theta$.
+\\
+Entropy Sampling selects instances with the highest entropy to query for labeling, focusing on the most ambiguous cases.
+It measures the amount of information needed to encode a distribution and is often used as a measure of uncertainty.
+\begin{align}
+	\mathcal{H}(\mathbf{x})=-\sum_{i=0}^{9} P_\theta(y_i \mid \mathbf{x}) \log P_\theta(y_i \mid \mathbf{x})
+\end{align}
+Where $y_i$ stands for the probability to predict class $i$. \\
+Vote Entropy Sampling is a query strategy designed for ensemble models.
+It calculates the ensemble predictions, then calculates the probability distribution over the predictions which are used to determine the Entropy $\mathcal{H}$.
+Vote Entropy then selects instances with the highest Entropy.
+\\
+Margin Sampling selects the instances where the difference between the most likely and second most likely class is the smallest, indicating high ambiguity.
+\begin{align}
+	Margin(\mathbf{x})=P_\theta\left(\hat{y}_1 \mid \mathbf{x}\right)-P_\theta\left(\hat{y}_2 \mid \mathbf{x}\right)
+\end{align}
+where $\hat{y}_1$ and $\hat{y}_2$ are the most likely and the second likely predicted labels given by $\hat{y}_1=\operatorname{arg}\underset{y}{\operatorname{max}} \  p_\theta(y \mid x)$ and $\hat{y}_2=\operatorname{arg}\underset{y \neq \hat{y}_1}{\operatorname{max}} \  p_\theta(y \mid x)$, respectively.
+\\
+Ranked batch-mode Sampling~\cite{ranked_batch_mode_query_strategy} weights uncertainty and diversity.
+It is designed for querying multiple samples, as traditional approaches such as uncertainty sampling may add redundant samples when querying multiple samples at once.
+\\
+\begin{align}
+	RBM(x)&=\alpha \left(1-\Phi\left(x, X_{\text {labeled }}\right)\right) \notag \\ &+(1-\alpha) \ Uncertainty(x)
+\end{align}
+Where $\alpha=\frac{\left|X_{\text {unlabeled}}\right|}{\left|X_{\text {unlabeled}}\right|+\left|X_{\text {labeled}}\right|}$, $X_{\text {labeled}}$ and $X_{\text {unlabeled}}$ are the labeled dataset and unlabeled pool, respectively.
+$\Phi$ denotes the similarity given by the Cosine Similarity metric.
+\begin{figure*}[!h]
+	\begin{center}
+		\includegraphics[width=0.95\linewidth]{fig/test_accuracy.pdf}
+	\end{center}
+	\caption{Test accuracy versus the number of annotated samples that were added to the train dataset by querying instances.
+As examples are queried and added to the training dataset, the model is retrained, and its performance is evaluated on the test set.}
+	\label{accuracy_plot}
+\end{figure*}
+\begin{figure*}[!h]
+	\begin{center}
+		\includegraphics[width=0.95\linewidth]{fig/conf_mat_margin_sampling.pdf}
+	\end{center}
+	\caption{Confusion Matrix of models trained with Margin Sampling.}
+	\label{conf_mat_margin_sampling}
+\end{figure*}
+\subsection{Evaluation metrics}
+In the following section, multiple evaluation metrics are used to gain insights into understanding the differences and quality of each query method and the single-model vs. ensemble-model approach. \\
+Due to the fact, that MNIST is class-balanced, a well-working metric is the accuracy. It is defined as as the number of correctly classified instances divided by the total number of instances. \\
+Furthermore, confusion matrices are created to evaluate systemic errors in the prediction of certain numbers.  
 \section{Results}\label{sec:results}
-Quantitative results: Tables or plots comparing the query strategies across scenarios. Also maybe key perforamnce differences (accuracy vs number of labeled examples)
+A Logistic Regression model that is trained on the whole labeled train dataset of MNIST reaches a test accuracy of $92.58\%$.
+This serves as a theoretical maximum that can be reached given a sufficient amount of train data.
+\\
+By plotting the accuracy on the test samples vs.
+the number of queried examples in figure~\ref{accuracy_plot}, convergence behaviour of different query methods can be analyzed.
+On MNIST, Margin Sampling reaches the highest Accuracy with 92.08\% for both models and converges the fastest, followed by Random Sampling.
+Entropy Sampling and Uncertainty Sampling perform worse than Random Sampling.
+Only after querying 5000 additional samples, the difference in accuracy between both strategies and Random Sampling is 0.25\%.
+For most query strategies, performance between a single model and an ensemble of models does not differ.
+\\
+Furthermore, it can be seen that ensemble models converge faster than single models with the same query method, with small train datasets, but differences between models shrink with an increased number of queried samples.
+\\
+The ensemble that is trained with Vote Entropy sampling performs better than the models trained with Entropy sampling with an accuracy difference of 2.08\% to the single model and 1.79\% to the ensemble, indicating that query strategies specifically adapted for ensemble models can boost classification performance.
+\\
+Figure~\ref{conf_mat_margin_sampling} and shows the confusion matrices of the best performing query strategy, Margin Sampling.
+The single model trained with Margin Sampling achieves the lowest accuracy of $86\%$ when an input image contains the number 5, whereas the ensemble reaches has 1\% more accuracy in this case, which also resembles the worst number for the ensemble.
+In both cases, the most likely prediction is 3 when the number 5 is shown.
 
 \section{Discussion}\label{sec:discussion}
-Qualitative analysis: Why certain strategies perform better or worse in specific settings. Trends between simple models and complex models
+Results show that Margin Sampling achieves the highest accuracy and fastest convergence by prioritizing instances with the greatest ambiguity, effectively adding diversity to the training dataset.
+It does not show high systemic errors.
+In contrast, Uncertainty and Entropy Sampling underperform compared to Random Sampling, possibly due to redundant sample selection.
 \\
+Across query strategies, the ensemble converges faster than single models the smaller train datasets are, leveraging prediction diversity in early stages.
+However, its advantage diminishes as the labeled train dataset grows.
+This leads to the conclusion that ensembles help to reduce the impact of single model biases in small datasets.
+For most strategies, the ensemble's added diversity does not translate into substantial performance gains, as shown by its marginal improvement over single models.
+Both, single and ensemble models, converge similarly for the same sampling strategy, indicating that a single Logistic Regression Model is able to effectively learn MNIST when given enough data.
+Furthermore, early biases can be reduced by using an ensemble.
+Using more complex models and data might lead to the ensemble converging to a higher accuracy than single models, as is often seen.
 \\
-Impact of initial dataset size
-\\
-\\
-Most methods require knowledge about the balance of the dataset and might be disadvantageous to use in case of a disbalanced dataset.
-In real world scenarios, this is not known.
-Both used datasets are balanced so the impact of query strategies on imbalanced datasets was not tested.
-\\
-Only image-based datasets were tested. Not vector- or text-based.
+This study uses a logistic regression model and a simple, balanced dataset, which simplifies the analysis but limits generalizability.
+While the findings highlight the effectiveness of certain strategies for shallow models, it is unclear if these results hold for more complex models or imbalanced datasets.
+Future work should explore diverse data types, like text, tabular or time series data, complex models, and imbalanced datasets to better understand the broader applicability of these findings and see whether complex models show similar convergence profiles.
 
 \section{Conclusion}\label{sec:conclusion}
-Summary of findings and insights (maybe that a strategy works best with small datasets while another one is robust on bigger ones)
+This report systematically evaluated active learning query strategies using logistic regression on MNIST.
+Margin Sampling is shown to be the most effective strategy, achieving the highest accuracy and fastest convergence by focusing on instances with the greatest ambiguity.
+Ensemble models demonstrated faster convergence in early training stages, possibly due to higher diversity but offered limited performance gains over single models as labeled datasets grew.
 \\
-Implications: practical recommendations?
+Key findings suggest that ensembles are particularly advantageous when dealing with small datasets, reducing biases from single-model predictions.
+However, since single logistic regression models already perform well on MNIST when there’s enough labeled data, using an ensemble doesn’t add much benefit in this case.
+The underperformance of uncertainty- and entropy-based strategies compared to Random Sampling indicates that redundant sample selection remains a challenge, as Random Sampling may unintentionally add variety to the data.
 \\
-Future work? (More diverse datasets, exploring additional strategies, applying in real-world tasks)
-
-\newpage
-
-\section{Cool ideas}
-\begin{itemize}
-	\item Show the typical framework of active learning in a figure in the introduction
-	\item Accuracy vs. fraction of data size in relation to the datset (percentage of used data points)
-\end{itemize}
-
-\section{Good formulations}
-From the active learning survey: Entropy is an information-theoretic
-measure that represents the amount of information needed to “encode” a distri-
-bution. As such, it is often thought of as a measure of uncertainty or impurity in machine learning.
-
-\section{Comparison to the other papers}
-% FROM HERE THIS IS NOT GOING TO BE PART OF THE FINAL PART
-Previous papers do not: Is everything here now mentioned previously? if yes, then it can be deleted.
-\begin{itemize}
-	\item \cite{schröder_revisitinguncertaintybasedquerystrategies}: Does not compare across different sizes of the pool, and for simple models.
-	\item \cite{comparativesurveydeepactive}: Does not compare impact of different initial train set size and strategies on shallow classifiers.
-
-	\item According to \cite{ueno_benchmarkingofquerystrategies}: \textcolor{red}{check if true:} Does not compare different sizes of the initial starting sets as well as how simple and more complex perform with the query methods
-
-	\item \cite{werner_comparableactivelearning}: They focus on a wide range of domains, this paper only compares methods in the image domain.
-I compare strategies for a simple logistic regression model and a complex CNN across different starting sizes of the annotated dataset.
-\\
-\end{itemize}
-% TO HERE
+Future research should extend these insights to more complex models and datasets, including imbalanced and multi-domain scenarios such as text and tabular data.
+Additionally, exploring novel query strategies tailored to diverse use cases could further optimize active learning pipelines for real-world applications.
 

### 2025-02-03 code/utils.py
Add utility methods

@@ -1,19 +1,22 @@
 import os, random, pickle
 import time
 
-from modAL import ActiveLearner
-from modAL.uncertainty import classifier_margin
+from modAL import ActiveLearner, Committee
+from modAL.uncertainty import classifier_margin, classifier_uncertainty
 from sklearn.datasets import fetch_openml
-from sklearn.metrics import accuracy_score, log_loss
+from sklearn.metrics import accuracy_score, log_loss, pairwise_distances
 from sklearn.linear_model import LogisticRegression
 import numpy as np
 import tensorflow as tf
 import torch
 from sklearn.model_selection import train_test_split
 from skorch import NeuralNetClassifier
+from skorch.callbacks import GradientNormClipping
 from torch import nn
 from cnn_model import CNN
 
+is_cnn=False
+device="cpu"
 
 def initialize_random_number_generators(seed):
     np.random.seed(seed)
@@ -39,71 +42,38 @@ def load_MNIST(random_seed, validation_split=0.2):
     X_train_full, y_train_full = X[:60000], y[:60000]  # This is the default way of extracting train and test data
     X_test, y_test = X[60000:], y[60000:]
 
-    # Create validation split
-    X_train, X_val, y_train, y_val = train_test_split(
-        X_train_full, y_train_full, test_size=validation_split, random_state=random_seed
-    )
+    if validation_split == 0.0:
+        X_train = X_train_full
+        y_train = y_train_full
+        X_val, y_val = [], []
+    else:
+        # Create validation split
+        X_train, X_val, y_train, y_val = train_test_split(
+            X_train_full, y_train_full, test_size=validation_split, random_state=random_seed
+        )
 
     return X_train, y_train, X_test, y_test, X_val, y_val, X, y
 
 
-def load_CIFAR():
-    (X_train, y_train), (X_test, y_test) = tf.keras.datasets.cifar10.load_data()
-
-    # Normalize pixel values as they are rgb values between 0 and 255
-    X_train = X_train.astype('float32') / 255.0
-    X_test = X_test.astype('float32') / 255.0
-
-    # Transpose the datasets to have the channel (3) first and then height (1) and width (2)
-    X_train = np.transpose(X_train, (0, 3, 1, 2))
-    X_test = np.transpose(X_test, (0, 3, 1, 2))
-
-    # Convert labels to 1D array
-    y_train, y_test = y_train.flatten(), y_test.flatten()
-
-    X_whole = np.concatenate((X_train, X_test), axis=0)
[…]

### 2025-02-03 code/utils.py
Final push to utility methods

@@ -15,10 +15,25 @@ from skorch.callbacks import GradientNormClipping
 from torch import nn
 from cnn_model import CNN
 
-is_cnn=False
-device="cpu"
+is_cnn = False
+device = "cpu"
+
 
 def initialize_random_number_generators(seed):
+    """
+    Initializes random number generators for NumPy, Python's random module, and PyTorch
+    to ensure reproducibility.
+
+    Args:
+        seed (int): The seed value to initialize the random number generators.
+
+    Notes:
+        - Sets the seed for NumPy's random number generator.
+        - Sets the seed for Python's built-in random module.
+        - Sets the seed for PyTorch's random number generator (both CPU and CUDA).
+        - deterministic behavior in PyTorch by setting `torch.backends.cudnn.deterministic = True`.
+        - Disables CuDNN benchmarking (`torch.backends.cudnn.benchmark = False`)
+    """
     np.random.seed(seed)
     random.seed(seed)
     # Cuda should not be used but I found it hard to produce reproducible results with
@@ -30,6 +45,27 @@ def initialize_random_number_generators(seed):
 
 
 def load_MNIST(random_seed, validation_split=0.2):
+    """
+    Loads the MNIST dataset from OpenML, normalizes the pixel values, and optionally splits
+    the training data into training and validation sets.
+
+    Args:
+        random_seed (int): The seed for random number generation to ensure reproducibility
+                           when splitting the dataset.
+        validation_split (float, optional): The proportion of the training data to be used
+                                            as validation data. Defaults to 0.2.
+
+    Returns:
+        tuple: A tuple containing:
+            - X_train (numpy.ndarray): Training set features.
+            - y_train (numpy.ndarray): Training set labels.
+            - X_test (numpy.ndarray): Test set features.
+            - y_test (numpy.ndarray): Test set labels.
+            - X_val (numpy.ndarray): Validation set features (empty if validation_split=0.0).
+            - y_val (numpy.ndarray): Validation set labels (empty if validation_split=0.0).
+            - X (numpy.ndarray): Full dataset features.
+            - y (numpy.ndarray): Full dataset labels.
+    """
     mnist = fetch_openml('mnist_784', version=1, parser="auto")
 
     X, y = mnist.data.to_numpy(), mnist.target.to_numpy()
@@ -55,9 +91,24 @@ def load_MNIST(random_seed, validation_split=0.2):
[…]

### 2025-02-03 doc/paper.tex
Add final push with updated paper figure and little addition to method section

@@
 \subsection{Query Strategies}
-A multitude of query strategies are used to compare a single model to an ensemble of models. As a control metric, Random Sampling is used to give a baseline for other query strategies.\\
+A multitude of uncertainty-based query strategies are used to compare a single model to an ensemble of models.
+Uncertainty-based methods are chosen because they prioritize samples the models are least confident about, making them an effective approach in active learning.
+They try to improve model performance with fewer labeled samples by focusing on unclear cases, which are particularly relevant in a setting where labeling is expensive.
+Additionally, they are computationally straightforward.
+\\
+As a control metric, Random Sampling is used to give a baseline for other query strategies. \\
 Uncertainty Sampling selects the least certain instances for labeling.
